Comparing corrosion control treatments for drinking water using a robust Bayesian generalized additive model

### 2022-07-29 Rmarkdown/paper-si.Rmd
added fig s9, response to reviewer 3
remove fig 9
fig 7 annotations, response doc

This document contains 8 figures, 4 tables, and 12 pages.

@@

>>>>

This document contains 9 figures, 4 tables, and 13 pages.

@@

```{r figure-s9, fig.cap="**Figure S9. (a)** True colour in pipe loop feedwater, by month and year. **(b)** Aluminum in feedwater, by orthophosphate dose. "}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-s9.png"))
```

### 2022-07-29 Rmarkdown/paper.Rmd
fig 7 annotations, response doc

Non-linear time trends and seasonal variation can be difficult to describe without over- or under-fitting. They can also make it difficult to identify a stable phase, as typical guidance for analyzing pipe loop study data suggests [@wysock_statistical_1995]. The most common methods for describing pipe loop data are usually inadequate to describe complex nonlinear variation [@simpson_modelling_2018].

@@

Censoring typically arises in pipe loop studies when analyte concentrations fall below the detection limit (i.e., left-censoring). Left-censoring is often addressed by substituting a constant for values below the detection limit and sometimes by removing the censored values. Neither approach is advisable as both introduce artificial patterns that can distort results, sometimes seriously [@helsel_statistics_2012]. Popular statistical tests (e.g., the paired t-test and its generalization, the one-way repeated measures ANOVA) cannot accommodate censoring [@helsel_statistics_2012]. The non-parametric rank-sum test does allow for left-censoring at one detection limit, but it is not adequate for more complex censoring schemes (e.g., multiple detection limits or interval censoring). Imputation methods such as regression on order statistics may be useful [@helsel_statistics_2012], but in a time series the censored data may be correlated, and imputation in the context of a full data model is usually preferable [@hopke_multiple_2001].

@@

Here, we propose a Bayesian generalized additive model for pipe loop study data that is sufficiently flexible for most pipe loop datasets. The collection of measurement series representing variation in a given parameter over time---often lead--- is represented as the sum of five components (Figure 1b):

1. An average representing each pipe (all other components are centered on zero).
2. A smooth seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature (this component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it).
3. A smooth global trend describing multi-year variation in the parameter of interest across all pipes.
4. A set of smooth pipe-specific trends that describe the deviations from the global multi-year trend particular to each pipe.
5. An autocorrelated error term.

@@

An advantage of our approach is the ability to move beyond point estimates and intervals: Bayesian models yield the full posterior distribution of all parameters, and all quantities derived from those parameters, including the predicted values. This allows for full propagation of uncertainty through all stages of post-processing and analysis [@beck_multi-scale_2022]. Moreover, the use of additive smoothing splines permits a description of time series using multiple nonlinear components, which can yield a more nuanced understanding of differences in lead concentrations over time.

@@

The continuous-time autocorrelated error term was successful in capturing residual first-order autocorrelation, estimated in Figure 5b using the Spearman rank correlation coefficient as a measure that is robust against outliers. Since the difference between model predictions and the censoring limit is not an accurate estimate of residual model error, we simulated residuals for model assessment by imputing censored values with posterior predictions and computing the residuals using the augmented dataset. Simulated residuals exhibited minimal autocorrelation, while predictions that excluded the autocorrelation term yielded much stronger residual autocorrelation. The distribution of the autocorrelation coefficient for each model, $\phi$, is shown in Figure 5c, and simulated residuals as a function of time are shown in Figure S8.

@@

Over much of the study, the dissolved fraction (<0.45 µm) dominated total lead release (Figure 6). During the conditioning period in particular---before orthophosphate was introduced---particles represented a minor contribution. Orthophosphate treatment caused a sharp decrease in dissolved lead (<0.45 µm) and a concomitant increase in the particulate fraction, which is consistent with previous observations and geochemical models [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017; @ng_new_2018; @li_controlling_2020]. Given the prominent seasonality in dissolved lead release, the particulate fraction exhibited inverse seasonality, with higher particulate fractions in winter (Figure 6). The increase in particulate lead in the final weeks of the study period is not easily explained, but it may be related to increasing feedwater aluminum concentrations (Figure S9)---previous work has shown that aluminum and orthophosphate interact to increase particulate lead release [@trueman_seasonal_2022].

@@

Effluent lead concentrations in pipe loop studies typically exhibit a pronounced negative trend after commissioning, followed by relative stability for the remainder of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant. This general pattern is represented in our model by the global multi-year trend (Figure 4b), which excludes pipe-specific and seasonal water quality variation. The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term. Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a). Both fractions exhibited positive and negative trends in subsequent years that are difficult to explain, but they may be a response to nonseasonal changes in water quality over the study (see *Trends in particulate lead release*).

@@

The variation in dissolved lead captured in Figure 8a neglects seasonality by design. And while the seasonal component of the model (Figure 8b) is described as a function of time (ordinal day), this is a proxy for the physical and chemical inputs that are expected to have caused it. The true seasonal relationship is probably complex and difficult to determine from the study data, but temperature and influent iron concentration are plausible drivers: both varied seasonally and have been shown to impact lead release (Figure 8b--c) [@masters_increased_2015; @masters_seasonal_2016; @trueman_galvanic_2017; @trueman_understanding_2016]. And while dissolved organic carbon has also been shown to impact lead release [@korshin_influence_2005; @lin_release_2008; @king_role_2022], true colour---a proxy for dissolved organic carbon [@redden_chemical_2021]---was not strongly seasonal (Figure S9).

@@

Given that pipe loop studies are expected to remain important for drinking water quality management, time series analysis will be  necessary for reliable statistical analysis of the data they generate. Methods that do not account for the effect of repeated measurements---and the reduced information content that results from autocorrelation---are likely to be less useful, and may even be misleading. Methods should be flexible enough to account for nonlinear time trends, and they should be robust against extreme values. And since the goal of pipe loop studies is often to minimize the release of lead and other metals into drinking water, concentrations may fall below the analytical range for a given analyte. The most useful statistical methods, then, will be well-suited to modeling censored data.

>>>>

Non-linear time trends and seasonal variation can be difficult to describe without over- or under-fitting. They can also make it difficult to identify a stable phase, as typical guidance for analyzing pipe loop study data suggests [@wysock_statistical_1995]. The most common models for pipe loop data are usually inadequate to describe complex nonlinear variation [@simpson_modelling_2018].

@@

Censoring typically arises in pipe loop studies when analyte concentrations fall below the detection limit (i.e., left-censoring). Left-censoring is often addressed by substituting a constant for values below the detection limit and sometimes by removing the censored values. Neither approach is advisable as both introduce artificial patterns that can distort results, sometimes seriously [@helsel_statistics_2012]. Popular statistical tests---for example, the paired t-test and its generalization, the one-way repeated measures ANOVA---cannot accommodate censoring [@helsel_statistics_2012]. The non-parametric rank-sum test does allow for left-censoring at one detection limit, but it is not adequate for more complex censoring schemes (e.g., multiple detection limits or interval censoring). Imputation methods such as regression on order statistics may be useful [@helsel_statistics_2012], but in a time series the censored data may be correlated, and imputation in the context of a full data model is usually preferable [@hopke_multiple_2001].

@@

Here, we propose a Bayesian generalized additive model for pipe loop study data that is sufficiently flexible for most pipe loop datasets. The collection of measurement series representing variation in a given parameter over time---often lead---is represented as the sum of five components (Figure 1b):

1. An average representing each pipe (all other components are centered on zero).
2. A smooth seasonal trend (i.e., a function of ordinal day), whose strength depends on the magnitude of the seasonal variation in water quality and temperature. This component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it.
3. A smooth global trend describing multi-year variation in the parameter of interest across all pipes (i.e., a function of time).
4. A set of smooth pipe-specific trends that describe the deviations from the global multi-year trend particular to each pipe.
5. An autocorrelated error term.

@@

An advantage of our approach is the ability to move beyond point estimates and intervals: Bayesian models yield the full posterior distribution of all parameters and all quantities derived from those parameters, including the predicted values. This allows for propagation of uncertainty through all stages of post-processing and analysis [@beck_multi-scale_2022]. Moreover, the use of additive smoothing splines permits a description of time series using multiple nonlinear components, which can yield a more nuanced understanding of differences in analyte concentrations over time.

@@

The continuous-time autocorrelated error term was successful in capturing residual first-order autocorrelation, estimated in Figure 5b using the Spearman rank correlation coefficient as a measure that is robust against outliers. Since the difference between model predictions and the censoring limit is not an accurate estimate of residual model error, we simulated residuals for model assessment by imputing censored values with posterior predictions and computing the residuals using the augmented dataset. Simulated residuals were minimally autocorrelated, while predictions that excluded the autocorrelation term yielded much stronger residual autocorrelation. The distribution of the autocorrelation coefficient for each model, $\phi$, is shown in Figure 5c, and simulated residuals as a function of time are shown in Figure S8.

@@

Over much of the study, the dissolved fraction (<0.45 µm) dominated total lead release (Figure 6). During the conditioning period in particular---before orthophosphate was introduced---particles represented a minor contribution. Orthophosphate treatment caused a sharp decrease in dissolved lead (<0.45 µm) and a concomitant increase in the particulate fraction, which is consistent with previous observations and geochemical models [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017; @ng_new_2018; @li_controlling_2020]. Given the prominent seasonality in dissolved lead release, the particulate fraction was inversely seasonal, with higher particulate fractions in winter (Figure 6). The increase in particulate lead in the final weeks of the study is not easily explained, but it may be related to increasing feedwater aluminum concentrations (Figure S9)---previous work has shown that aluminum and orthophosphate interact to increase particulate lead release [@trueman_seasonal_2022].

@@

Effluent lead concentrations in pipe loop studies often decrease sharply after commissioning, followed by relative stability for the remainder of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant. This general pattern is represented in our model by the global multi-year trend (Figure 4b), which excludes pipe-specific and seasonal water quality variation. The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term. Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a). The subtler subsequent positive and negative trends in both fractions are difficult to explain, but they may be a response to nonseasonal changes in water quality (see *Trends in particulate lead release*).

@@

The variation in dissolved lead captured in Figure 8a neglects seasonality by design. And while the seasonal component of the model (Figure 8b) is described as a function of time (ordinal day), this is a proxy for the physical and chemical inputs that are expected to have caused it. The true seasonal relationship is probably complex and difficult to determine from the study data, but temperature and influent iron concentration are plausible drivers: both varied seasonally and have been shown to impact lead release (Figure 8b--c) [@masters_increased_2015; @masters_seasonal_2016; @trueman_galvanic_2017; @trueman_understanding_2016]. And although dissolved organic carbon has also been shown to impact lead release [@korshin_influence_2005; @lin_release_2008; @king_role_2022], true colour---a proxy for dissolved organic carbon [@redden_chemical_2021]---was not strongly seasonal (Figure S9).

@@

Given that pipe loop studies are expected to remain important for drinking water quality management, time series methods will be  necessary for reliable statistical analysis of the data they generate. Methods that do not account for the effect of repeated measurements---and the reduced information content that results from autocorrelation---are likely to be less useful, and may even be misleading. Methods should be flexible enough to account for nonlinear time trends, and they should be robust against extreme values. And since the goal of pipe loop studies is often to minimize the release of lead and other metals into drinking water, concentrations may fall below the analytical range for a given analyte. The most useful statistical methods, then, will be well-suited to modeling censored data.
